# Stream Nutrient Concenrations Before and After Hurricane Hugo

## Background

Following Hurricane Hugo, streams in Puerto Rico experienced prolonged changes in nutrient concentration.  Schaefer et al. (2000) documented this change by analyzing data from four long term study sites in streams in Bisely, Puerto Rico.  This project seeks to replicate their analysis.

## Data

We accessed stream nutrient data from the four study sites in Bisley, Puerto Rico through the Environmental Data Initiative and [McDowel and International Institute of Tropical Forestry (IJTF) 2024](https://doi.org/10.6073/PASTA/F31349BEBDC304F758718F4798D25458). 

The four stream sites include Quebrada one-Bisly (Q1/BQ1) Chemistry Data, Quebrada two-Bisley (Q2/BQ2) Chemistry Data, Quebrada three-Bisley (Q3/BQ3) Chemistry Data, and Puente Roto Mameyes (MPR/PRM) Chemistry Data.

## Methods

We narrowed data down to the seven-year study window from 1988 through 1994 and isolated the five analyzed nutrients: calcium (Ca), potassium (K), magnesium (Mg), nitrate (NO3-N), and ammonia (NH4-N).  We created a 9 week moving average for each nutrient at each study site, plotting them against time in the below figures to compare trends in nutrient levels before and after the hurricane in 1998 (indicated by the vertical dashed line).

## Results
Reproduction of Figure 3 from Schaefer et al. (2000)
```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| fig-alt: "Figure of concentrations of calcium, potassium, magnesium, nitrate, and ammonia in four streams in Bisley, Puerto Rico before and after the September 1989 hurricane Hugo.  Concentrations of calcium, potatssium, and nitrate spike immediate after the hurricane. Calcium and potasium remain elevated relative to pre-hurricane levels for serveral years following the hurricane for multiple sites."
library(tidyverse)
#Set readable labels for nutrients
nutrient_labels <- c(
  k_mgl = "K mg/L",
  ca_mgl = "Ca mg/L",
  mg_mgl = "Mg mg/L",
  nh4_n_ugl = "NH4-N ug/L",
  no3_n_ugl = "NO3-N ug/L"
)
#Call the cleaned data file with moving averages created by 1_clean_data.R, housed in output
four_stream_avg_plot <- read_csv("../output/four_stream_avg_long.csv")
#Plot concentration against time, separating by site
ggplot(
  data = four_stream_avg_plot,
  mapping = aes(
    x = window_start,
    y = Concentration,
    color = sample_id,
    linetype = sample_id
  )
) +
  geom_line() +
  #Separate nutrients into individual facets
  facet_grid(
    Nutrient ~ .,
    scales = "free",
    switch = "y",
    labeller = labeller(Nutrient = nutrient_labels),
    axes = "all",
    axis.labels = "all_y",
  ) +
  #Clean up the plot formatting
  labs(color = "Site", linetype = "Site") +
  scale_y_continuous(name = "Concentration") +
  scale_x_date(name = "Year") +
  scale_color_grey() +
  geom_vline(xintercept = as.Date("1989-09-01"), linetype = "dashed") +
  theme(
    panel.background = element_blank(),
    strip.background = element_blank(),
    strip.placement = "outside",
    axis.ticks = element_line(color = "black"),
    axis.minor.ticks.x.top = element_line(),
    axis.minor.ticks.x.bottom = element_line(),
    axis.minor.ticks.y.left = element_line(),
    axis.line = element_line(color = "black")
  )
```
 Full color version:
```{r, echo=FALSE, message=FALSE, warning=FALSE}
#| fig-alt: "Figure of concentrations of calcium, potassium, magnesium, nitrate, and ammonia in four streams in Bisley, Puerto Rico before and after the September 1989 hurricane Hugo.  Concentrations of calcium, potatssium, and nitrate spike immediate after the hurricane. Calcium and potasium remain elevated relative to pre-hurricane levels for serveral years following the hurricane for multiple sites."
library(tidyverse)
#Set readable labels for nutrients
nutrient_labels <- c(
  k_mgl = "K mg/L",
  ca_mgl = "Ca mg/L",
  mg_mgl = "Mg mg/L",
  nh4_n_ugl = "NH4-N ug/L",
  no3_n_ugl = "NO3-N ug/L"
)
#Call the cleaned data file with moving averages created by 1_clean_data.R, housed in output
four_stream_avg_plot <- read_csv("../output/four_stream_avg_long.csv")
#Plot concentration against time, separating by site
ggplot(
  data = four_stream_avg_plot,
  mapping = aes(
    x = window_start,
    y = Concentration,
    color = sample_id,
    linetype = sample_id
  )
) +
  geom_line() +
  #Separate nutrients into individual facets
  facet_grid(
    Nutrient ~ .,
    scales = "free",
    switch = "y",
    labeller = labeller(Nutrient = nutrient_labels),
    axes = "all",
    axis.labels = "all_y",
  ) +
  #Clean up the plot formatting
  labs(color = "Site", linetype = "Site") +
  scale_y_continuous(name = "Concentration") +
  scale_x_date(name = "Year") +
  geom_vline(xintercept = as.Date("1989-09-01"), linetype = "dashed") +
  theme(
    panel.background = element_blank(),
    strip.background = element_blank(),
    strip.placement = "outside",
    axis.ticks = element_line(color = "black"),
    axis.minor.ticks.x.top = element_line(),
    axis.minor.ticks.x.bottom = element_line(),
    axis.minor.ticks.y.left = element_line(),
    axis.line = element_line(color = "black")
  )
```

## References
 McDowell, William H., and USDA Forest Service. International Institute Of Tropical Forestry (IITF). 2024. “Chemistry of Stream Water from the Luquillo Mountains.” Environmental Data Initiative. [https://doi.org/10.6073/PASTA/F31349BEBDC304F758718F4798D25458].

Schaefer, Douglas. A., William H. McDowell, Fredrick N. Scatena, and Clyde E. Asbury. 2000. “Effects of Hurricane Disturbance on Stream Water Concentrations and Fluxes in Eight Tropical Forest Watersheds of the Luquillo Experimental Forest, Puerto Rico.” Journal of Tropical Ecology 16 (2): 189–207. [https://doi.org/10.1017/s0266467400001358]. 